Pool Model - A unified Mathematical\\ Modeling Framework for Food Science

### 2025-11-28 paper/pool_model.tex
replace \mu with \omega in all equations

    L &\stackrel{\mu}{\longrightarrow} D\\

@@

    G &\stackrel{\mu'}{\longrightarrow} D

@@

    \dot{L} &= -(\mu + \lambda) L\\
    \dot{G} &= \lambda L + \alpha G - \mu' G\\
    \dot{D} &= \mu  L + \mu' G 

@@

        \dot{L} &= -(\mu + \lambda) L\\
        \dot{G} &= \lambda L + \alpha \tilde{R} G-\mu' G\\
        \dot{D} &= \mu  L + \mu' G\\

@@

        \dot{L} &= -(\mu + \lambda) L\\
        \dot{G} &= \lambda L + \alpha G\left(1-\frac{L+G+D}{N_t}\right)-\mu' G\\
        \dot{D} &= \mu  L + \mu' G.

@@

        All curves were calculated using the model (\ref{eq:ode_3pool_resource2}) with parameters $\alpha=2.5~\mathrm{s}^{-1}$, $\mu'=0.5~\mathrm{s}^{-1}$, $N_t = 10^4$.
        The blue curve shows a significant dip in the behavior of the system where the death of the bacteria is present $\mu = 0.5~\mathrm{s}^{-1}$.
        Other curves show the case of no bacteria dying $\mu=0~\mathrm{s}^{-1}$ and represent different lengths of lag phases: short $\lambda=10^{-2}~\mathrm{s}^{-1}$ (dark green), intermediate $\lambda=10^{-3}~\mathrm{s}^{-1}$ (light green) and long one $\lambda=10^{-4}~\mathrm{s}^{-1}$ (yellow). Note that the steady state for $N(t)$ and $D(t)$ is for this model $\lim_{t\to\infty} N(t)=0$ and $\lim_{t\to\infty} D(t)=N_t$, because of the depletion of the resource pool.

@@

Moreover, for $\mu'=0$ the equations for $L$ and $D$ can be readily integrated resulting in:
        L(t) &= N_0 e^{-(\mu+\lambda)t}\\
        D(t) &= N_0 \mu \frac{1-e^{-(\mu+\lambda)t}}{\mu+\lambda

@@

For constant parameters $\mu$ and $\lambda$ the integrals for $L$ and $D$ can be inserted into the differential equation for $G$.

@@

For the functions $L$ and $D$ it holds: $\lim_{t\to\infty} L(t) = 0$ and  $\lim_{t\to\infty} D(t) =  N_0\frac{\mu}{\mu+\lambda} $.
It follows immediately that $\lim_{t\to\infty} G(t) = N_t-N_0\frac{\mu}{\mu+\lambda} $. The lag time $t_L$ can be approximated by the time at which half of the initial population is in the growth phase, i.e., $N_0=2G(t_L)$.

@@

    1  &= 2\lambda e^{\alpha t_L}\frac{1-e^{-(\mu+\lambda+\alpha)t_L}}{\mu+\lambda+\alpha}.
For $\alpha > \mu+\lambda$  we can approximate $t_L$ by:
    t_L &\approx -\frac{1}{\alpha}\ln\left(\frac{2\lambda}{\mu+\lambda+\alpha}\right).

@@

        \dot{L} &=\gamma(t) G - (\mu + \lambda) L\\

@@

        \dot{D} &= \mu  L

@@

        The other model parameters are $\alpha=2.5~\mathrm{s}^{-1}$, $\mu'=0.5~\mathrm{s}^{-1}$, $\beta=5~\mathrm{s}^{-1}$, $\lambda=10^{-2}~\mathrm{s}^{-1}$, $\mu = 0~\mathrm{s}^{-1}$ for each curve.

@@

We assume that the toxin causes the death of the species $A$ with rate $\mu$.

@@

        \dot{G}_A &= \lambda_A R L_A + \alpha_A R G_A - \frac{\mu}{\kappa N_t} T G_A\\

@@

        The toxin $T_B$ causes the cell death of bacteria $A$ with rate constant $\mu = 0.2~\mathrm{s}^{-1}$.

@@

% ================================================================================================
\section{Spatial effects}
As described above, the proposed model operates with the total number of the bacterial counts for
different species and states in the system, not accounting for their spatial distributions.
Assuming that the bacterial system is uniformly distributed (sometimes also called 'well-stirred')
we expect the \ac{ode} description to be valid.
However, in reality spatial effects can play a significant role for population survival forcing some
species to develop competitive mechanisms such as cell adhesion to surfaces in the areas rich with
nutrients~\citep{htuson_bacteriasurface_2013}.
Another mechanism is the motility of bacteria, e.g., the flagellates of \textit{E. coli} or
\textit{B. subtilis} which are capable of chemotactic motion driven by oxygen
concentrations~\citep{decoene_microscopic_2011}.
The bacterial social interactions such as release of toxins or public goods can also
significantly affect the spatial structure of the community~\citep{blanchard_bacterial_2015}.
These effects can only be treated within a spatial framework that is able to describe the observed
effects.\\
Compared to other modeling techniques such as \acp{pde}, \acp{abm}~\citep{nagarajan_agent-based_2022-1} force researchers to think in
terms of cellular processes, thus bridging the gap from a
collective (bulk) model to an individual-based (single-cell) description.
Our goal is to provide an \ac{abm} which can be used to derive the given \ac{ode} model as a
coarse-grained case.
In this way, we can show that the bulk-based \ac{ode} model is directly related to the mechanistic
single-cell behavior of the bacteria.
To this end, we consider the system consisting of two bacterial species that can be described by the two internal states: lag phase and growth.
Both of these species grow using the same resource pool $R$ but with different growth constants
$\alpha_{A/B}$ and different lag phase transition rates $\lambda_{A/B}$.
Moreover, species $A$ is able to release a certain antimicrobial component $I$ that shows an
inhibitory effect on species $B$ (amensalism).
We assume that $L,G$ are measured in terms of the overall volume and $R,I$ are measured as total
amount.
The \acp{ode} system of the bacterial population can be written as:
%
%
\begin{align}
    \begin{split}
        \dot{L}_A &= -\lambda_A R L_A\\
        \dot{G}_A &= \lambda_A R L_A + \alpha_A R G_A\\
        \dot{L}_B &= -\lambda_B R L_B\\
        \dot{G}_B &= \lambda_B R L_B + \frac{\alpha_B}{1 + \mu_I I} R G_B.
    \end{split}
    \label{eq:spatial_limit_F}
\end{align}
%
Moreover, the microenvironment of our system contains the resource and the inhibitor molecules
\begin{align}
    \begin{split}
        \dot{R} &= -\frac{\alpha_A}{N_t} R G_A-\frac{\alpha_B}{(1 + \mu_I I) N_t} R G_B \\
        \dot{I} &= \kappa G_A.
    \end{split}
    \label{eq:spatial_limit_H}
\end{align}

@@

    \dot{G}_B &= \frac{\alpha_B}{1 + \mu_I I} R G_B + \lambda_B R L_B\\
    \dot{R} &= -\frac{\alpha_A}{N_t} R G_A-\frac{\alpha_B}{(1 + \mu_I I) N_t} R G_B\\

@@

        \mu_I &= \frac{\nu_I}{V_D}\\

@@

% ================================================================================================
\newpage
\beginsupplement
% ================================================================================================
\section{General mathematical formalism}
\label{ssec:supplement1}
Consider the system consisting of $N$ species in $M$ different states.
In the general representation, all discussed in the main part systems  can be described with the Ordinary Differential Equations (\acp{ode}) using a non-linear function $\mbF$:
\begin{equation}
   \dot{\mbx} = \mbF(\mbx, \mbm, \mbp, \mbu),
\label{eq:model_ODE_x}
\end{equation}
where  $\mbx \in \mathbb{R}^{N}  \otimes \mathbb{R}^{M}$ is a vector describing bacterial count.
In matrix form it can be written as
\begin{equation}
    \mbx = \begin{pmatrix}
        x^1_1  & \dots & x^1_N  \\
        \vdots &       & \vdots \\
        x^M_1  & \dots & x^M_N  \\
            \end{pmatrix}
    \label{eq:model_bact}
\end{equation}
with $x_{j}^{i}$ describing the bacterial counts for species $i$ in state $j$.
$\mbu$ is the vector of different inputs of the model, e.g., temperature, pressure etc.
The vector $\mbm \in \mathbb{R}^{K}$ includes all variables corresponding to the microenvironment of the system which can be concentration of the inhibitors or activators:
\begin{equation}
    \dot{\mbm} = \mbH (\mbx, \mbm, \mbp, \mbu).
    \label{eq:model_microenv}
\end{equation}
Here function $\mbH$ is in general case also non-linear.\\
%
The advantage of the following formulation is its generality.
Depending on the knowledge about the system as well as the needed level of accuracy one can define an arbitrary number of states $M$ and control the complexity of the model.
Moreover, opposite to the already known Predictive Microbiology like Baranyi and Roberts one can easily change.\\
%
To calculate the bacterial count for one species $i$, we simply need to sum the tensor/vector $\mbx$ over all possible states:
\begin{equation}
    N_i = \sum_k \mbx_i^k.
\label{eq:N_pro_species}
\end{equation}
While the total bacterial count of all species of our system is
\begin{equation}
    N = \sum_i \sum_k \mbx_i^k.
\label{eq:N_total}
\end{equation}
%
%
%
Using such matrix representation of the bacterial system, we can rewrite, for example, the three-pool model with limiting resource described with equations~\ref{eq:ode_3pool_resource} in a more compact form.
The concentrations of the pools are then the components of our variable vector $\mbx (t) = (L(t), G(t), D(t))^T$.
%
\begin{align}
    \dot \mbx  &= \mbF(\mbx, \mbm, \mbp, \mbu) = \begin{pmatrix}
        -(\mu + \lambda) & 0               & 0      \\
        \lambda          & \alpha R - \mu' & 0      \\
        \mu              & \mu'            & -\beta 
    \end{pmatrix} \mbx\\
    \dot \mbm &= \mbH (\mbx, \mbm, \mbp, \mbu) = -\frac{\alpha}{N_t} R G.
\end{align}
Separating linear and non-linear parts in two terms yields
\begin{equation}
    \dot \mbx = \begin{pmatrix}
        -(\mu + \lambda) & 0       & 0 \\
         \lambda         & \alpha  & 0 \\
         \mu &  \mu'& - \beta 
    \end{pmatrix} 
    \mbx + \frac{\alpha}{N_t} x^2 \sum_i^3 x^i \begin{pmatrix} 0 \\ 1 \\ 0  \end{pmatrix}.
\label{eq:x_repres_3pools_resource2}
\end{equation}
Here the first term is the linear part and the second term define the quadratic behavior of the system.
%
\paragraph{Generalized Lotka-Volterra}

@@

        Inhibition                  & $\mu_I$       & $1.0\cdot 10^{-7}$~\unit{\per\micro\metre\cubed}

>>>>

    L &\stackrel{\omega}{\longrightarrow} D\\

@@

    G &\stackrel{\omega'}{\longrightarrow} D

@@

    \dot{L} &= -(\omega + \lambda) L\\
    \dot{G} &= \lambda L + \alpha G - \omega' G\\
    \dot{D} &= \omega  L + \omega' G

@@

        \dot{L} &= -(\omega + \lambda) L\\
        \dot{G} &= \lambda L + \alpha \tilde{R} G-\omega' G\\
        \dot{D} &= \omega  L + \omega' G\\

@@

        \dot{L} &= -(\omega + \lambda) L\\
        \dot{G} &= \lambda L + \alpha G\left(1-\frac{L+G+D}{N_t}\right)-\omega' G\\
        \dot{D} &= \omega  L + \omega' G.

@@

        All curves were calculated using the model (\ref{eq:ode_3pool_resource2}) with parameters $\alpha=2.5~\mathrm{s}^{-1}$, $\omega'=0.1~\mathrm{s}^{-1}$, $N_t = 10^4$.
        The blue curve includes cell-death of the bacteria in the lag phase $\omega = 0.15~\mathrm{s}^{-1}$ and thus exhibits a small dip before its exponential growth.
        Other curves show the case of no bacteria dying $\omega=0~\mathrm{s}^{-1}$ and represent different lengths of lag phases: short $\lambda=10^{-2}~\mathrm{s}^{-1}$ (dark green), intermediate $\lambda=10^{-3}~\mathrm{s}^{-1}$ (light green) and long one $\lambda=10^{-4}~\mathrm{s}^{-1}$ (yellow). Note that the steady state for $N(t)$ and $D(t)$ is for this model $\lim_{t\to\infty} N(t)=0$ and $\lim_{t\to\infty} D(t)=N_t$, because of the depletion of the resource pool.

@@

Moreover, for $\omega'=0$ the equations for $L$ and $D$ can be readily integrated resulting in:
        L(t) &= N_0 e^{-(\omega+\lambda)t}\\
        D(t) &= N_0 \omega \frac{1-e^{-(\omega+\lambda)t}}{\omega+\lambda

@@

For constant parameters $\omega$ and $\lambda$ the integrals for $L$ and $D$ can be inserted into the differential equation for $G$.

@@

For the functions $L$ and $D$ it holds: $\lim_{t\to\infty} L(t) = 0$ and  $\lim_{t\to\infty} D(t) =  N_0\frac{\omega}{\omega+\lambda} $.
It follows immediately that $\lim_{t\to\infty} G(t) = N_t-N_0\frac{\omega}{\omega+\lambda} $. The lag time $t_L$ can be approximated by the time at which half of the initial population is in the growth phase, i.e., $N_0=2G(t_L)$.

@@

    1  &= 2\lambda e^{\alpha t_L}\frac{1-e^{-(\omega+\lambda+\alpha)t_L}}{\omega+\lambda+\alpha}.
For $\alpha > \omega+\lambda$  we can approximate $t_L$ by:
    t_L &\approx -\frac{1}{\alpha}\ln\left(\frac{2\lambda}{\omega+\lambda+\alpha}\right).

@@

        \dot{L} &=\gamma(t) G - (\omega + \lambda) L\\

@@

        \dot{D} &= \omega  L

@@

        The other model parameters are $\alpha=2.5~\mathrm{s}^{-1}$, $\omega'=0.5~\mathrm{s}^{-1}$, $\beta=5~\mathrm{s}^{-1}$, $\lambda=10^{-2}~\mathrm{s}^{-1}$, $\omega = 0~\mathrm{s}^{-1}$ for each curve.

@@

We assume that the toxin causes the death of the species $A$ with rate $\omega$.

@@

        \dot{G}_A &= \lambda_A R L_A + \alpha_A R G_A - \frac{\omega}{\kappa N_t} T G_A\\

@@

        The toxin $T_B$ causes the cell death of bacteria $A$ with rate constant $\omega = 0.2~\mathrm{s}^{-1}$.

@@

% ================================================================================================
\section{Spatial effects}
As described above, the proposed model operates with the total number of the bacterial counts for
different species and states in the system, not accounting for their spatial distributions.
Assuming that the bacterial system is uniformly distributed (sometimes also called 'well-stirred')
we expect the \ac{ode} description to be valid.
However, in reality spatial effects can play a significant role for population survival forcing some
species to develop competitive mechanisms such as cell adhesion to surfaces in the areas rich with
nutrients~\citep{htuson_bacteriasurface_2013}.
Another mechanism is the motility of bacteria, e.g., the flagellates of \textit{E. coli} or
\textit{B. subtilis} which are capable of chemotactic motion driven by oxygen
concentrations~\citep{decoene_microscopic_2011}.
The bacterial social interactions such as release of toxins or public goods can also
significantly affect the spatial structure of the community~\citep{blanchard_bacterial_2015}.
These effects can only be treated within a spatial framework that is able to describe the observed
effects.\\
Compared to other modeling techniques such as \acp{pde}, \acp{abm}~\citep{nagarajan_agent-based_2022-1} force researchers to think in
terms of cellular processes, thus bridging the gap from a
collective (bulk) model to an individual-based (single-cell) description.
Our goal is to provide an \ac{abm} which can be used to derive the given \ac{ode} model as a
coarse-grained case.
In this way, we can show that the bulk-based \ac{ode} model is directly related to the mechanistic
single-cell behavior of the bacteria.
To this end, we consider the system consisting of two bacterial species that can be described by the two internal states: lag phase and growth.
Both of these species grow using the same resource pool $R$ but with different growth constants
$\alpha_{A/B}$ and different lag phase transition rates $\lambda_{A/B}$.
Moreover, species $A$ is able to release a certain antimicrobial component $I$ that shows an
inhibitory effect on species $B$ (amensalism).
We assume that $L,G$ are measured in terms of the overall volume and $R,I$ are measured as total
amount.
The \acp{ode} system of the bacterial population can be written as:
%
%
\begin{align}
    \begin{split}
        \dot{L}_A &= -\lambda_A R L_A\\
        \dot{G}_A &= \lambda_A R L_A + \alpha_A R G_A\\
        \dot{L}_B &= -\lambda_B R L_B\\
        \dot{G}_B &= \lambda_B R L_B + \frac{\alpha_B}{1 + \omega_I I} R G_B.
    \end{split}
    \label{eq:spatial_limit_F}
\end{align}
%
Moreover, the microenvironment of our system contains the resource and the inhibitor molecules
\begin{align}
    \begin{split}
        \dot{R} &= -\frac{\alpha_A}{N_t} R G_A-\frac{\alpha_B}{(1 + \omega_I I) N_t} R G_B \\
        \dot{I} &= \kappa G_A.
    \end{split}
    \label{eq:spatial_limit_H}
\end{align}

@@

    \dot{G}_B &= \frac{\alpha_B}{1 + \omega_I I} R G_B + \lambda_B R L_B\\
    \dot{R} &= -\frac{\alpha_A}{N_t} R G_A-\frac{\alpha_B}{(1 + \omega_I I) N_t} R G_B\\

@@

        \omega_I &= \frac{\nu_I}{V_D}\\

@@

% ================================================================================================
\newpage
\beginsupplement
% ================================================================================================
\section{General mathematical formalism}
\label{ssec:supplement1}
Consider the system consisting of $N$ species in $M$ different states.
In the general representation, all discussed in the main part systems  can be described with the Ordinary Differential Equations (\acp{ode}) using a non-linear function $\mbF$:
\begin{equation}
   \dot{\mbx} = \mbF(\mbx, \mbm, \mbp, \mbu),
\label{eq:model_ODE_x}
\end{equation}
where  $\mbx \in \mathbb{R}^{N}  \otimes \mathbb{R}^{M}$ is a vector describing bacterial count.
In matrix form it can be written as
\begin{equation}
    \mbx = \begin{pmatrix}
        x^1_1  & \dots & x^1_N  \\
        \vdots &       & \vdots \\
        x^M_1  & \dots & x^M_N  \\
            \end{pmatrix}
    \label{eq:model_bact}
\end{equation}
with $x_{j}^{i}$ describing the bacterial counts for species $i$ in state $j$.
$\mbu$ is the vector of different inputs of the model, e.g., temperature, pressure etc.
The vector $\mbm \in \mathbb{R}^{K}$ includes all variables corresponding to the microenvironment of the system which can be concentration of the inhibitors or activators:
\begin{equation}
    \dot{\mbm} = \mbH (\mbx, \mbm, \mbp, \mbu).
    \label{eq:model_microenv}
\end{equation}
Here function $\mbH$ is in general case also non-linear.\\
%
The advantage of the following formulation is its generality.
Depending on the knowledge about the system as well as the needed level of accuracy one can define an arbitrary number of states $M$ and control the complexity of the model.
Moreover, opposite to the already known Predictive Microbiology like Baranyi and Roberts one can easily change.\\
%
To calculate the bacterial count for one species $i$, we simply need to sum the tensor/vector $\mbx$ over all possible states:
\begin{equation}
    N_i = \sum_k \mbx_i^k.
\label{eq:N_pro_species}
\end{equation}
While the total bacterial count of all species of our system is
\begin{equation}
    N = \sum_i \sum_k \mbx_i^k.
\label{eq:N_total}
\end{equation}
%
%
%
Using such matrix representation of the bacterial system, we can rewrite, for example, the three-pool model with limiting resource described with equations~\ref{eq:ode_3pool_resource} in a more compact form.
The concentrations of the pools are then the components of our variable vector $\mbx (t) = (L(t), G(t), D(t))^T$.
%
\begin{align}
    \dot \mbx  &= \mbF(\mbx, \mbm, \mbp, \mbu) = \begin{pmatrix}
        -(\omega + \lambda) & 0               & 0      \\
        \lambda          & \alpha R - \omega' & 0      \\
        \omega              & \omega'            & -\beta
    \end{pmatrix} \mbx\\
    \dot \mbm &= \mbH (\mbx, \mbm, \mbp, \mbu) = -\frac{\alpha}{N_t} R G.
\end{align}
Separating linear and non-linear parts in two terms yields
\begin{equation}
    \dot \mbx = \begin{pmatrix}
        -(\omega + \lambda) & 0       & 0 \\
         \lambda         & \alpha  & 0 \\
         \omega &  \omega'& - \beta
    \end{pmatrix} 
    \mbx + \frac{\alpha}{N_t} x^2 \sum_i^3 x^i \begin{pmatrix} 0 \\ 1 \\ 0  \end{pmatrix}.
\label{eq:x_repres_3pools_resource2}
\end{equation}
Here the first term is the linear part and the second term define the quadratic behavior of the system.
%
\paragraph{Generalized Lotka-Volterra}

@@

        Inhibition                  & $\omega_I$    & $1.0\cdot 10^{-7}$~\unit{\per\micro\metre\cubed}
